---
title: 'MITx 12.340x Global Warming Science: Final Exam'
author: "Michael Hunt"
date: "May 8, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Q1

Write down an expression for "F", the total radiative energy flux per unit area from an ideal blackbody radiator, as a function of its absolute temperature "T" and the Stefan-Boltzmann constant "sigma". Explicitly indicate any mathematical operations with the appropriate symbol (e.g. '+', '-', '*', '/', '^').

Answer: 

$$F=\sigma T^4$$

### Q2

Which one of these scientists is generally credited as being the first to postulate that our atmosphere provides a greenhouse effect?

 Isaac Newton  
 Jean Baptiste Fourier  -- correct
 John Tyndall  
 Albert Einstein
 
 Answer:
 
In the 1820s, Fourier estimated the temperature the Earth should have in the absence of atmospheric effects and found it was much too cold. He speculated that heat trapping by the atmosphere might be responsible for the relative warmth of the surface.

### Q2

What is the main reason that summer is generally warmer than winter?

 The Earth is closer to the sun in summer.  
 Without snow on the ground, the albedo is lower in summer.  
 Sunlight incident on the summer hemisphere is more nearly perpendicular to the Earth's surface, so that more sunlight is absorbed per unit area.  
 The atmosphere has a greater concentration of water vapor in summer.
 
Answer:  

Summer is warmer than winter principally because the obliquity of the earth's rotation axis with respect to the orbital plane means that sunlight strikes the summer hemisphere at a smaller angle to the vertical than it does the winter hemisphere.

### Q3  

What is the main reason that summer is generally warmer than winter?

 The Earth is closer to the sun in summer.  
 Without snow on the ground, the albedo is lower in summer.  
 Sunlight incident on the summer hemisphere is more nearly perpendicular to the Earth's surface, so that more sunlight is absorbed per unit area. - correct  
 The atmosphere has a greater concentration of water vapor in summer.
 
Answer:  
Summer is warmer than winter principally because the obliquity of the earth's rotation axis with respect to the orbital plane means that sunlight strikes the summer hemisphere at a smaller angle to the vertical than it does the winter hemisphere.

### Q4

Which one of these scientists is generally credited as being the first to discover which atmospheric gases are responsible for the greenhouse effect?

 John Tyndall  -- correct
 Svante Arrhenius  
 Max Planck  
 James Maxwell


Answer:  

John Tyndall devised an apparatus for measuring infrared absorption by various gases, and with it he discovered that most infrared absorption in our atmosphere is owing to water vapor, carbon dioxide and a few other trace gases.

### Q5

Write down an expression for the average amount of sunlight absorbed by the Earth per unit surface area, Fs, as a function of the solar constant "S" and the planetary albedo "alpha". Explicitly indicate any mathematical operations with the appropriate symbol (e.g. '+', '-', '*', '/', '^').

$$F_s=\dfrac{S}{4}*(1-\alpha)$$


### Q6

To the nearest whole number, what is the ratio of the average amount of infrared back-radiation from the atmosphere and clouds absorbed by the Earth's surface, to the average amount of sunlight absorbed by the Earth's surface?

2

On average, the Earth's surface absorbs $168 \textrm{ W m}^{-2}$ of solar radiation and $324 \textrm{ W m}^{-2}$ of infrared back-radiation from the atmosphere and clouds.

### Q7

Most of what we usually call "weather" occurs in which of these layers of the atmosphere?

 The thermosphere.  
 he mesosphere.  
 The stratosphere.  
 The troposphere.  - correct
 
Answer:  

Almost all weather occurs in the troposphere, a region of the atmosphere that extends from the Earth's surface to the tropopause, which is usually between 8 and 14 km altitude. The great static stability of the stratosphere, which lies above the tropopause, strongly inhibits upward transport of water vapor and other transient gases and aerosols from the troposphere.

### Q8

On average, which one of these processes is primarily responsible for cooling the Earth's surface, to balance incoming absorbed sunlight?

 _Net_ infrared radiation  
 Turbulent flux of sensible heat away from the surface  
 Evaporation from the surface< - correct  
 Ocean currents
 
On average, evaporation of water from the surface carries away $78 textrm{ W m}^{-2}$, net infrared radiation carries away $66 \textrm{ W m}^{-2}$, and the sensible heat flux is about $24 \textrm{ W m}^{-2}$ .

### Q9  

Which one of these scientists is generally credited for being the first to estimate the average surface warming that would occur if the carbon dioxide content of the atmosphere were doubled?

 Jean Baptiste Fourier  
 John Tyndall  
 Svante Arrhenius - correct  
 Jule Charney
 
### Q10

Physically, why are the greenhouse gases, like water vapor and carbon dioxide, so much more effective in absorbing solar and/or infrared radiation than are oxygen and nitrogen?

 The concentrations of greenhouses gases in the atmosphere are greater than the concentrations of oxygen and nitrogen.  

 Greenhouse gases have many more degrees of freedom, such as rotational and vibrational transitions, in interacting with radiation than do nitrogen and oxygen, owing to the greater complexity of their molecular structure. - correct  
 The molecular weights of greenhouse gases are greater than those of oxygen and nitrogen.  The larger physical size of more complex molecules means that they can interact more readily with infrared radiation.
 
Answer:  
Molecular oxygen and nitrogen are simple homonuclear diatomic molecules with no electric dipole moment, while the greenhouse gases have rich absorption spectra owing to numerous vibrational, rotational, and rotational-vibrational modes.

### Q11

Which of these is responsible for most of the Earth's albedo?

 Clouds -- correct
 Sea ice  
 Land ice  
 Deserts
 
Clouds are responsible for more than 70% of the Earth's total albedo.

### Q12

In a dry atmosphere, which one of these expresses a condition for instability to convection?

 Temperature decreases with altitude.  
 Temperature increases with altitude.  
 Pressure decreases with altitude.  
 Entropy decreases with altitude.
 
### Q19

In terms of its effect on annual mean surface air temperature, which of these greenhouse gases has the greatest aggregate effect?

 Nitrous oxide  
 Carbon dioxide  
 Ozone  
 Water 
 vapor - correct
 
Answer:

Mostly because of its much greater abundance, water vapor dominates other greenhouse gases in its effect on surface temperature.

### Q20

It has been suggested that Mars once had liquid water on its surface. If so, this was probably mainly because the Martian atmosphere likely had...

 more clouds than it does presently.  
 less oxygen than it does presently.  
 more carbon dioxide than it does presently. - correct  
 fewer particles than it does presently.
 
Answer:  

It is possible that the Martian atmosphere had more carbon dioxide early in its history, warming the surface and thereby allowing for the presence of liquid water. (a thicker atmosphere)
 
### Q21

What is the average residence time of a molecule of water in the atmosphere?

 3 days  
 2 weeks - correct  
 1 year  
 20 years 
 
Answer: 

Water is constantly evaporating from the surface and falling back to the surface as rain or snow. The average residence time of a water molecule in the atmosphere is about 2 weeks.

### Q22

Which one of these processes is generally thought to be primarily responsible for the timing of the glacial-interglacial cycles over the past few million years?  

Variations in earth's orbit around the sun and in the tilt of its rotation axis - correct  
Variations in solar radiative output  
Changes in the greenhouse gas content of the atmosphere  
Changes in ocean circulation  

Answer:  

The Milankovitch cycles of the obliquity of the Earth's rotation axis, its precession, and the ellipticity of the Earth's orbit around the sun are generally recognized as the root cause of the glacial-interglacial cycles of the past few million years, but changes in greenhouse content and ocean circulation are thought to be important feedbacks in this mode of climate change.
 
### Q23  

Which one of these statements best describes the effect of clouds on the mean surface air temperature of Earth?

 By reflecting sunlight, clouds always cool the surface.  
 By absorbing and re-radiating infrared radiation, clouds always warm the surface.  
 Depending on their altitude and optical properties, clouds can cool or warm the surface. - correct  
 Clouds have only a very small effect on surface temperature.  
 
Answer:  

Low-level clouds emit infrared radiation at temperatures not very different from those of the surface, so they have little greenhouse effect, but their high albedo yields a strong cooling. High clouds emit at temperatures much lower than those of the Earth's surface, and so are effective greenhouse agents, though they also reflect some sunlight. Thus, depending in their altitude and optical properties, clouds can cool or warm the surface.

### Q24

Comprehensive calculations of radiative equilibrium yield surface temperatures that are somewhat warmer than observed, and tropopause temperatures that are somewhat colder than observed. What is the main reason for this?

 Radiative equilibrium neglects the effects of large-scale circulation of the atmosphere and ocean.  
 Radiative equilibrium calculations specify cloudiness rather than including it as part of the solution.  
 Radiative equilibrium calculations often neglect aerosols.  
Dry and moist convection are neglected.</text> - correct  

Answer:  

Convection transports enthalpy upward and thereby cools the lower atmosphere and surface while warming the middle and upper troposphere.

### Q25

What is the principal cause of the large-scale circulation of the ocean?

 Lunar and solar tides  
 Surface winds   -- correct
 Horizontal gradients of ocean temperature</text>
 Horizontal gradients of salinity
 
Answer:  
 
Surface winds are the primary drivers of ocean circulation. On large scales, the curl of the surface wind stress drives Ekman pumping, which in turn drives large-scale ocean circulations, and small scale wind storms may contribute to vertical mixing, which also drives large-scale overturning circulations in the oceans.

### Q26

Which of these statements best describes Sverdrup balance in the oceans?

 A balance between pressure gradients and Coriolis accelerations.  
 A balance between the vertical pressure gradient and gravity.  
 A balance between the torque produced by the curl of the surface wind stress and advection of planetary vorticity by meridional flow of water. - correct  
 A balance between Coriolis acceleration and meridional advection of zonal momentum.  

Answer:  

The Sverdrup relation entails a balance between the curl of the surface wind stress and the meridional advection of planetary vorticity. It yields a good first-order description of the large-scale horizontal flow of the oceans.

### Q28

Over the next 100 years, which of the reservoirs listed below is predicted to play the largest role in removing anthropogenic CO2 from the atmosphere?

 Rocks/subsurface (including fossil fuels)  
 Plants  
 Ocean - correct  
 Soils

Answer:  

Roughly half of the anthropogenic carbon emitted to the atmosphere over the next century will wind up in the oceans.

### Q29

Where will the greatest amount of anthropogenic CO2 end up over the next 1 million years?

 In soils  
 As fossil fuels  
 Dissolved in the ocean  
 As calcium carbonate in seafloor sediments - correct
 
Answer:  

On the time scale of a million years, most of the carbon we emit to the atmosphere, after dissolving in the ocean, will wind up as calcium carbonate deposits on the seafloor.

### Q30

Which of the following best describes why Hadley circulations are limited to the Earth's tropics?

 Outside the tropics, Coriolis accelerations are strong enough to balance temperature gradients (through the thermal wind relation). - correct  
 It is too cold outside the tropics.  
 There is not enough water vapor outside the tropics.  
 The lack of eddy activity outside Earth's tropics prevents overturning circulations from developing.  
 
Answer:  

Hadley circulations occur where there is not enough angular momentum available to create wind speeds necessary so that Coriolis accelerations may balance horizontal pressure gradient accelerations associated with horizontal temperature gradients. Outside the tropics, there is more than enough angular momentum available to satisfy this balance.

### Q31  

Which of these atmospheric phenomena is primarily responsible for meridional atmospheric transport of heat in middle and high latitudes?

 The Hadley Circulation  
 Moist convection  
 The Ferrel Cell  
 Stationary planetary waves excited by air flow over mountains 
 Travelling baroclinic waves with typical horizontal dimensions of a few thousand kilometers - correct  
 
Answer:  

Outside the tropics, baroclinic eddies, associated with "highs" and "lows" on weather maps, are responsible for most of the poleward enthalpy transport in the atmosphere.
 
### Q32

Satellite and weather balloon observations of the upper stratosphere show that it has been cooling over the past few decades, even though the troposphere has been warming. Which of these trends best explains this observation?

 Decreasing ozone content of the stratosphere  
 Increasing carbon dioxide concentration of the atmosphere (including the stratosphere) - correct   
 Decreasing amplitude of the 11-year solar cycle    
 Changing aerosol content of the stratosphere  

Answer: 

Both increasing carbon dioxide concentrations and decreasing ozone concentrations result in a cooling of the stratosphere. However, the cooling of the upper stratosphere is primarily driven by increasing radiative cooling due to carbon dioxide, while in the lower stratosphere the effect of ozone is important.


 
### Q33

Theoretically, which best describes the dependence of equilibrium surface temperature on atmospheric carbon dioxide concentration?

 Equilibrium temperature increases linearly with CO2 content.  
 Equilibrium temperature increases quadratically with CO2 content.  
 Equilibrium temperature increases logarithmically with CO2 content. -- correct  Equilibrium temperature increases exponentially with CO2 content.
 
See this [article in Skeptical Science](http://www.skepticalscience.com/C02-emissions-vs-Temperature-growth.html)

Using data from that article:

```{r}
CumEmissions=c(0,500,1000,1500,2000,2500) # GtC
inAtmosFraction=0.45 # fraction remainiong in atmosphere
leftInAtmos=inAtmosFraction*CumEmissions # GtC
GtCper_ppmRise=2.12 # additional GtC in atmosphere required for 1 ppm rise in CO2 concentration
C0 =280 # pre-industrial CO2 concentration in atmosphere in ppm
CO2conc=C0 + leftInAtmos/GtCper_ppmRise

RF=function(x){
    5.35*log(x)
}

RFCO2 = 5.35*RF(CO2conc/C0) # radiative forcing depends logarithmically on CO2 concentration


RFdoubling=RF(2)
sensitivity=3 # temperature rise in K for a doubling of CO2
Tanomaly=sensitivity*RFCO2/RFdoubling
plot(log(CO2conc),Tanomaly)
fit=lm(Tanomaly~log(CO2conc))
abline(fit$coefficients[1],fit$coefficients[2])
```
Primarily because of the exponential tails of the pressure-broadened line spectra of carbon dioxide, increasing carbon dioxide concentration in the atmosphere results in an increase in radiative forcing that varies nearly logarithmically with concentration.