---
title: "MITx12.340x GWS"
author: "Michael Hunt"
date: "February 27, 2016"
output: html_document
---

### A little climate physics

#### Problem 2: Equator to pole temperature difference

The annual mean incident solar radiation $I(\phi)$ as a function of latitude $\phi$ is given approximately by

$$I(\phi) = 342 \left(1 - 0.48 \dfrac{(3 \sin^2\phi -1)}{2}\right)$$

```{r}
I<-function(phi){
    342*(1 - 0.48 *(1/2)*(3 *(sin(phi))^2 -1))
}
I(0)
I(pi/2)


albedo<-function(phi){
    albedo<-0.3
}
```
Suppose the atmosphere is transparent to short-wave radiation, but opaque to long-wave radiation. In this case, Ta=Te.

Now,

$$\sigma T_e^4(\phi)=(1-\alpha(\phi))I(\phi)$$

so

$$T_e(\phi) = \left[\frac{(1-\alpha(\phi))I(\phi)}{\sigma}\right]^{1/4}$$





```{r}
Te<-function(phi){
    sigma<-5.670373e-8
    ((1-albedo(phi))*I(phi)/sigma)^0.25
}
```
Hence we can find the atmospheric temperature difference between the equator and pole in radiative equilibrium, 

$$\Delta T_a  = T_a(0^\circ{}) - T_a(90^\circ{})$$

```{r}
Te(0)
Te(pi/2)
Te(0)-Te(pi/2)
```

### The Paleocene-Eocene Thermal Maximum
The current rate of emission of carbon dioxide by humans through fossil fuel use is around 7 Pg of Carbon per year. Suppose during the entire PETM 10 000 Pg of carbon was released into the atmosphere. Assuming this release took 200 thousand years*, what is the ratio of the current emission of carbon through the burning of fossil fuels to the average rate of release of carbon during the PETM?

```{r}
7/(10000/200000)
```

*NOTE: The estimate used here for the timescale over which Carbon was input into the atmosphere during the PETM is based on the duration of the entire event. Modern paleo-climatic studies have suggested the Carbon impulse may have been much shorter in duration - 10,000 years or even shorter. Using this shorter timescale would give a ratio 20 times smaller.